Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-02-21 paper/conclusion.tex
Elisabeth's comments and some of Gus's
Gus's comments

In this paper, we demonstrated that the dispersion of velocities in the
direction of galactic latitude, \vb, can be used as an age proxy, by showing
that there is no strong evidence for mass-dependent heating in low-mass
\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
increase with decreasing mass.
Although {\it vertical} velocity, \vz, is the quantity that most strongly
traces time-dependent orbital heating in the disc of the Galaxy, most stars
with measured rotation periods do not yet have radial velocities, so we used
velocity in the direction of Galactic latitude, \vb\, as a proxy for \vz.
Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
\vz\ introduces an additional velocity dispersion, which increases with
increasing Galactic latitude.
For this reason we did not attempt to convert \vb\ dispersions into ages using
an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore assumed that \vb\ velocity dispersion can be used to accurately
rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
on average, older than a group of stars with a small velocity dispersion.

% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
the evolution of stellar rotation period as a function of effective
temperature and age.
We found that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-\teff\ relation for old stars.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We showed that the period-\teff\ relations change shape over time in a way
that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.

% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
We also found that the oldest stars in the \mct\ catalog are cooler than 4500
K, which suggests that lower-mass stars remain active for longer, allowing
their rotation periods to be measured at older ages.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
The velocity dispersions of stars increase smoothly across the rotation period
gap, indicating that the gap does not separate two distinct populations.
Finally, we used kinematics to reveal a population of synchronized binaries
with rotation periods less than around 10 days.

>>>>

In this paper, we demonstrated that the dispersion of velocities in the
direction of galactic latitude, \vb, can be used as an age proxy, by showing
that there is no strong evidence for mass-dependent heating in low-mass
\kepler\ dwarfs: the velocity dispersions of K and M dwarfs, whose
main-sequence lifetimes are longer than around 11 Gyrs, do not appear to
increase with decreasing mass.
Although {\it vertical} velocity, \vz, is a quantity that has been
demonstrated to trace time-dependent orbital heating in the disc of the
Galaxy, most stars with measured rotation periods do not yet have radial
velocities, so we used velocity in the direction of Galactic latitude, \vb\,
as a proxy for \vz.
Using stars in the GUMS simulation, we showed that using \vb\ as a proxy for
\vz\ introduces an additional velocity dispersion, which increases with
increasing Galactic latitude.
For this reason we did not attempt to convert \vb\ dispersions into ages using
an age-velocity dispersion relation.
However, after removing high-latitude ($b>15^\circ$) stars from the sample, we
confirmed that using \vb\ instead of \vz\ does not introduce any
mass-dependent velocity dispersion bias into the sample.
We therefore assumed that \vb\ velocity dispersion can be used to accurately
rank stars by age, \ie\ a group of stars with a large velocity dispersion is,
on average, older than a group of stars with a small velocity dispersion.

% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We used the \vb\ velocity dispersions of stars in the \mct\ catalog to explore
the evolution of stellar rotation period as a function of effective
temperature and age.
We found that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-\teff\ relation for old stars.
Instead we found that, at young ages, rotation period is anti-correlated with
\teff: cooler stars spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We showed that the period-\teff\ relations change shape over time in a way
that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.

% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
We also found that the oldest stars in the \mct\ catalog are cooler than 4500
K, which suggests that lower-mass stars remain active for longer, allowing
their rotation periods to be measured at older ages.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
The velocity dispersions of stars increase smoothly across the rotation period
gap, indicating that the gap does not separate two distinct populations.
Finally, we used kinematics to indicate that there is a population of
synchronized binaries with rotation periods less than around 10 days.

### 2020-02-21 paper/results.tex
Gus's comments

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K follow a braking law that changes over
time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period decreases with increasing
mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for K-dwarfs.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

@@

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperate for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Only stars cooler than 5000 K are plotted in
    the bottom panel in order to isolate populations above and below the
    period gap, which only extends up to temperatures of $\sim$4600 K.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

Figure \ref{fig:gap} shows the velocity dispersions of stars in the \mct\
sample, with stars subdivided into three groups: those that rotate more
quickly than the major rotation period distribution (green triangles), those
with rotation periods shorter than the gap (orange squares), and those with
rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology models, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
extends to temperatures as hot as 6000 K]{davenport2017}.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
On the other hand, the velocity dispersions of stars with rotation periods
shorter than the lower edge of the rotation period distribution (green
triangles) are not significantly smaller than the, presumed older,
orange-colored stars.
% This is an indication that many of these are synchronized binaries.
% The rotation periods of stars in synchronized binaries are tidally locked to
% the orbital period of the binary system and since synchronized binaries have
% short orbital periods this can produce old, seemingly isolated stars, with
% short rotation periods.
The large velocity dispersions of the most rapidly rotating stars indicates
that some fraction of these stars are old but rotating rapidly, and therefore
likely to be synchronized binaries.
% Given the strong correlation between binary fraction and primary mass
% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
% rotating) sequence, it is likely that
Figure \ref{fig:gap} indicates that there is an increased probability of stars
with rotation periods less than $\sim$10 days being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of visual binaries were rapid rotators, and the probability of a star
being a synchronized binary system substantially increased below rotation
periods of around 7 days \citep{simonian2019}.

>>>>

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period decreases with increasing
mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for K-dwarfs.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

@@

\begin{figure}
  \caption{
      Top: rotation period vs. effective temperate for stars in the \mct\
    sample, separated into three groups. Blue circles
      show stars with rotation periods longer than the
    period gap, orange squares show stars with rotation periods shorter than
    the gap, but longer than the lower edge of the main rotation period
    distribution, and green triangles show stars with rotation periods shorter
    than this lower edge.
    Stars were separated into these three groups using \citet{angus2019}
    gyrochronology models, with the scheme shown in the legend.
    Only stars cooler than 5000 K are plotted in
    the bottom panel in order to isolate populations above and below the
    period gap, which only extends up to temperatures of $\sim$4600 K.
    Bottom: the velocities of these groups of stars (in the direction of
    Galactic latitude, $b$) are shown as a function of rotation period.
    The black line indicates the velocity standard deviation as a function of
    period.
}
  \centering
    \includegraphics[width=1\textwidth]{gap}
\label{fig:gap}
\end{figure}

The bottom panel of figure \ref{fig:gap} shows the velocity dispersions of
stars in the \mct\ sample, with stars subdivided into three groups: those that
rotate more quickly than the major rotation period distribution (green
triangles), those with rotation periods shorter than the gap (orange squares),
and those with rotation periods longer than the gap (blue circles).
Stars were separated into these three groups using \citet{angus2019}
gyrochronology models, according to the scheme shown in the legend.
Only stars cooler than 5000 K are included in the bottom panel in order to
isolate populations above and below the period gap, which only extends up to a
temperature of $\sim$4600 K in our sample \citep[Although][found that the gap
extends to temperatures as hot as 6000 K]{davenport2017}.
In general, velocity dispersion increases with rotation period because both
quantities increase with age.
There is a smooth transition in velocity dispersion between stars with
rotation periods below and above the gap (orange squares to blue circles),
suggesting that these groups are part of the same galactic population.
Previously, only the overall velocity dispersions of all stars above and below
the gap have been compared, leading to the assumption that these groups belong
to two distinct populations \citep{mcquillan2014}.
On the other hand, the velocity dispersions of stars with rotation periods
shorter than the lower edge of the rotation period distribution (green
triangles) are not significantly smaller than the, presumed older,
orange-colored stars.
% This is an indication that many of these are synchronized binaries.
% The rotation periods of stars in synchronized binaries are tidally locked to
% the orbital period of the binary system and since synchronized binaries have
% short orbital periods this can produce old, seemingly isolated stars, with
% short rotation periods.
The large velocity dispersions of the most rapidly rotating stars indicates
that some fraction of these stars are old but rotating rapidly, and therefore
likely to be synchronized binaries.
% Given the strong correlation between binary fraction and primary mass
% \citep[\eg][]{price-whelan2020}, and the fact that M dwarfs take longer to
% transition from the `I` (rapidly rotating) sequence to the `C' (slowly
% rotating) sequence, it is likely that
Figure \ref{fig:gap} indicates that there is an increased probability of stars
with rotation periods less than $\sim$10 days being synchronized binaries.
This result is in agreement with a recent study which found that a large
fraction of visual binaries were rapid rotators, and the probability of a star
being a synchronized binary system substantially increased below rotation
periods of around 7 days \citep{simonian2019}.
